\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{bm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[round]{natbib}
\usepackage[colorlinks=true,allcolors=blue]{hyperref}
\usepackage{enumitem}

\newtheorem{definition}{Definition}
\newtheorem{assumption}{Assumption}

% Shorthands in the style of the companion papers.
\newcommand{\R}{R}                       % representation-valued random variable
\newcommand{\T}{T}                       % target / linguistic property
\newcommand{\rade}{\mathfrak{R}}         % Rademacher complexity
\newcommand{\hypclass}{\mathcal{H}}      % hypothesis class
\newcommand{\complexity}{c}              % complexity knob
\newcommand{\fit}{\mathrm{fit}}
\newcommand{\IIA}{\mathrm{IIA}}
\newcommand{\Exp}{\mathbb{E}}
\newcommand{\bigO}{\mathcal{O}}

\title{\textbf{The Complexity--Spuriousness Trade-off in Probing:\\
Toward a Unified Capacity Measure for Diagnostic\\ and Causal Interpretability}}

\author{Tiago Pimentel\\ \small ETH Z\"urich \\ \small \texttt{tiago.pimentel@inf.ethz.ch}}

\date{\today}

\begin{document}
\maketitle

\begin{abstract}
\noindent
Diagnostic probing and causal probing (via distributed alignment search, DAS) both
tend to find what one looks for if the probe is powerful enough.
\citet{pimentel2020information} show that a sufficiently powerful diagnostic probe
always recovers a target property, and \citet{sutter2025nonlinear} show that a
sufficiently complex alignment map makes any algorithm a distributed abstraction of
any network. Both are statements about a limit: at infinite complexity, either method
can fit any target, including a random one, so its reported score says nothing about
the network. Read at face value, the causal result says that causal probing is no
better than diagnostic probing---it leaves us where diagnostic probing did. But that
is a claim about the limit. We do not work at that limit: at finite complexity and
finite data, neither method can fit everything, and the extra structure in a causal
probe might make it more resistant to fitting \emph{spurious} targets, so causality
could help in practice even if both methods fail in the limit. This report proposes
measuring that directly---how well a method fits randomly relabelled data, which is a
version of its empirical Rademacher complexity---in a form that applies unchanged to
diagnostic probing and to DAS, and asks how this spurious-fitting capacity grows with
complexity for each. We describe
the measure and an experimental protocol on the hierarchical equality task, and report
preliminary results: on the correct labels, both a linear probe and a linear DAS map
recover the task's intermediate variable, so the feature is at once decodable and
interveneable. The comparison the report is about---how readily each fits
\emph{spurious} (randomly relabelled) targets as complexity grows---is the sweep we
are running.
\end{abstract}

\section{Motivation}\label{sec:motivation}

A probe is a supervised model trained to predict some property $\T$ from a
network's representations $\R$; if the probe succeeds, we are tempted to conclude
that the representations \emph{encode} the property
\citep{alain2017understanding,hewitt2019structural,belinkov2022probing}.
\citet{pimentel2020information} recast this practice information-theoretically:
probing estimates the mutual information $\mathrm{I}(\T;\R)$, and the
best estimate is obtained by the \emph{best-performing} probe one can train,
regardless of its complexity. The corollary is uncomfortable: under their
\autoref{sec:setup} assumptions, a contextual representation contains exactly as
much information about a word-level property as the underlying sentence does, so a
powerful enough probe recovers the property for reasons that need have nothing to
do with how---or whether---the network uses it.

Causal probing was meant to escape this critique by intervening rather than
merely decoding. Distributed alignment search \citep[DAS;][]{geiger2024finding}
learns a (possibly non-linear) alignment map $\phi$ onto the hidden space, performs
interchange interventions in the aligned basis, and measures the
\emph{interchange-intervention accuracy} (IIA) with which the network reproduces a
target algorithm's counterfactual behaviour.
\citet{sutter2025nonlinear} show that this escape is illusory: once the linearity
constraint on $\phi$ is lifted, \emph{any} algorithm can be aligned to \emph{any}
network with near-perfect IIA---even randomly initialised networks that cannot
solve the task. They call this the \textbf{non-linear representation dilemma}:
lifting the linearity constraint removes any principled way to trade off the
alignment map's complexity against its accuracy.

Read together, the two results are one failure seen twice: in each, an
interpretability method reports a score---probe accuracy, or IIA---chosen as the best
fit over a family of extraction functions, and in each, that score stops being
informative once the family is expressive enough. The practitioner has no principled
way to set the family's complexity.

There is a pessimistic reading of this, and it is worth stating plainly.
\citet{sutter2025nonlinear} show that in the limit of unconstrained alignment maps,
causal abstraction is vacuous in the same way diagnostic probing is: the method
succeeds regardless of what the network computes. At face value, this says that moving
from diagnostic to causal probing buys us nothing---we are left where
\citet{pimentel2020information} left us. But it is a claim about the limit, and it need
not hold in practice. At the finite map complexities people actually use, the extra
structure in a causal probe---it must act through the frozen network, and stay
consistent across several interventions---might make it harder to fit spurious
patterns than a diagnostic probe of comparable power. If so, causality does add
something: not in the limit, but in the regime we work in. Whether it does is an
empirical question. To answer it we need a measure of how readily each method fits
spurious targets that applies to both---a single notion of capacity, in the spirit of
Rademacher complexity \citep{bartlett2002rademacher} or VC dimension
\citep{vapnik1998statistical}---and a way to watch it grow with complexity.

\section{Setup and notation}\label{sec:setup}

We adopt notation compatible with the two companion papers. A network induces a
representation map $x \mapsto \R$, with $\R \in \mathbb{R}^{d}$. An
\textbf{interpretability method} $\mathcal{M}$ is characterised by three
ingredients: (i) a \emph{hypothesis class} $\hypclass_{\complexity}$ indexed by a
scalar \emph{complexity knob} $\complexity \in \mathbb{R}_{+}$; (ii) a per-example
\emph{fit} functional $\fit(h; \cdot) \in [0,1]$ measuring how well a hypothesis
$h \in \hypclass_{\complexity}$ reproduces a target on one example; and (iii) a
\emph{selection rule} that, given a finite sample $\mathcal{D}_n$ of size $n$,
returns $\hat h = \arg\max_{h \in \hypclass_{\complexity}}
\widehat{\fit}(h; \mathcal{D}_n)$ and reports the achieved fit
$s(\mathcal{M}) = \widehat{\fit}(\hat h; \mathcal{D}_n)$.

The two methods instantiate this template as follows.

\paragraph{Diagnostic probing.} $\hypclass_{\complexity}$ is a family of probes
$q_{\bm\theta}(t \mid \mathbf r)$---e.g.\ multi-layer perceptrons whose depth or
width grows with $\complexity$, as in \citet{pimentel2020information}. The targets
are property labels $t$, the inputs are activations $\mathbf r$, and
$\widehat{\fit}$ is accuracy (or, in the information-theoretic view, the negative
cross-entropy that lower-bounds $\mathrm{I}(\T;\R)$). The dataset is
$\mathcal{D}_n = \{(\mathbf r_i, t_i)\}_{i=1}^n$.

\paragraph{Causal probing (DAS).} $\hypclass_{\complexity}$ is a family of bijective
alignment maps $\phi$---e.g.\ reversible networks
\citep{gomez2017reversible} whose depth $L_{\mathrm{rn}}$ or hidden size
$d_{\mathrm{rn}}$ grows with $\complexity$, as in \citet{sutter2025nonlinear}. The
``label'' of an example is the counterfactual output the \emph{target algorithm}
prescribes, the inputs are (base, source) pairs, and $\widehat{\fit}$ is the IIA:
the rate at which the \emph{frozen} network, intervened upon in the aligned basis,
reproduces that output. The dataset is
$\mathcal{D}_n = \{(\mathbf x^{\mathrm{base}}_i, \mathbf x^{\mathrm{src}}_i,
\mathbf y^{\mathrm{cf}}_i)\}_{i=1}^n$.

The crucial structural difference---which we will argue is the heart of the
matter---is that for DAS the hypothesis is not a classifier of $\mathbf r$. It is a
\emph{reparametrisation} $\phi$ of the hidden space, after which a \emph{fixed}
intervention and the \emph{fixed} downstream network produce the prediction. The
induced labelling function is $x \mapsto \mathrm{N}\!\big(\mathrm{intervene}_\phi(x)\big)$,
and the family is constrained to invertible $\phi$.

\section{A unified capacity measure}\label{sec:measure}

\subsection{Spuriousness as fit to randomised targets}

We want to measure how much of a method's reported fit could be spurious---how well
it fits targets that carry no real signal about the network. The direct way to do
this is to relabel the data at random and check how well the method still fits it.
The expected best fit a hypothesis class achieves on such random targets is, up to
scaling, its empirical \emph{Rademacher complexity} \citep{bartlett2002rademacher};
we use it, in accuracy space, as our measure of spurious-fitting capacity.

\begin{definition}[Spurious-fitting capacity]\label{def:capacity}
Fix a representation sample of size $n$ and let $\bm\sigma$ denote a target
\emph{randomisation}: a relabelling of the examples that destroys any genuine
relationship between inputs and targets while preserving their marginals. The
spurious-fitting capacity of a method $\mathcal{M}$ at complexity $\complexity$ is
\begin{equation}\label{eq:capacity}
  \rade_n(\hypclass_{\complexity})
    \;\coloneqq\;
    \Exp_{\bm\sigma}\!\left[\,
      \sup_{h \in \hypclass_{\complexity}}
      \widehat{\fit}\big(h;\, \bm\sigma(\mathcal{D}_n)\big)
    \,\right] \;-\; s_{\mathrm{chance}},
\end{equation}
i.e.\ the expected best fit the class achieves on randomised targets, centred at
chance performance.
\end{definition}

\autoref{eq:capacity} is, up to a scaling constant and the usual move from a
margin loss to $0/1$ accuracy, the empirical Rademacher complexity of
$\hypclass_{\complexity}$ on the representation sample. Its appeal is threefold.
First, it is \emph{method-agnostic}: it specialises to diagnostic probing (random
token labels) and to DAS (a random target algorithm, or equivalently permuted
counterfactual outputs) without changing form. Second, it is exactly the
quantity practitioners already estimate: it is the control-task accuracy of
\citet{hewitt2019control}, and the random-network alignment of
\citet{sutter2025nonlinear} is its causal analogue.\footnote{We deliberately phrase
\autoref{def:capacity} in accuracy space rather than as the textbook
$\Exp_{\bm\sigma}[\sup_h \frac1n \sum_i \sigma_i h(\mathbf r_i)]$ so that it is
directly comparable across the two methods and directly estimable from the
existing code paths, which already train on randomised labels.} Third, it is
\emph{cheap}: a Monte-Carlo estimate requires only re-running the existing fitting
pipeline on permuted targets.

\subsection{The limit and the finite regime}\label{sec:two-modes}

\autoref{def:capacity} is not what the two theorems are about, and the difference
matters. The theorems are about a limit: at infinite complexity, both a probe and an
alignment map can fit any target on any amount of data, random targets included
\citep[Thm.~1]{pimentel2020information,sutter2025nonlinear}. In that limit the two
methods are equally uninformative and nothing tells them apart.

We never run at that limit. We fit at finite complexity on finite data, where neither
method can fit everything. There, the two can differ: a linear probe and a linear
rotation do not have the same ability to fit random relabellings, and that ability
grows differently as we make each more complex. \autoref{def:capacity} measures
exactly this ability, and its growth with complexity is what we study---not whether
the methods break in the limit (they do), but how their capacity to fit spurious
targets grows on the way there.

\subsection{What ``graceful'' should mean, and the comparability problem}

A natural temptation is to compare the two methods by plotting their capacity
against a shared notion of ``size''---say, parameter count---and asking whether one
grows like $\bigO(\sqrt{P})$ where the other grows like $\bigO(P)$. We caution
against taking the exponent itself as the contribution, for two reasons. First,
generic norm- or parameter-based bounds yield the \emph{same} $\bigO(\sqrt{P/n})$
form for any reasonable class \citep{neyshabur2015norm}, so an a priori difference
in exponent is not expected; a measured difference would have to come from
structure, not from counting. Second, ``depth of an MLP probe'' and ``depth of a
RevNet alignment map'' are not the same unit, so a raw $\complexity$-axis is not
commensurable.

Both problems are solved by using \emph{capacity itself} as the common axis.
Concretely, we propose to plot, for each complexity level, the \textbf{signal}
(fit on \emph{correct} targets) against the \textbf{spurious capacity} (fit on
\emph{randomised} targets, \autoref{def:capacity}). Each method traces a curve in
this \emph{signal--spurious plane} as $\complexity$ varies. A method
\textbf{fails gracefully} if its curve hugs the top-left---high signal recovered
per unit of spurious capacity spent---and \textbf{fails sharply} if signal and
spurious capacity rise together. This operationalises ``one method fails more
gracefully than the other'' without ever committing to incommensurable units, and
it directly generalises the selectivity of \citet{hewitt2019control} (which is the
single-point gap $\text{signal} - \text{spurious}$) into a \emph{scaling curve}.

\section{A mechanism for divergence}\label{sec:mechanism}

If the two methods do diverge in the signal--spurious plane, the explanation must
lie in structural constraints that one method imposes and the other does not. We
identify three constraints that are present in DAS but absent in diagnostic
probing, each of which should \emph{throttle} spurious capacity below what the raw
parameter count of $\phi$ would suggest:

\begin{enumerate}[leftmargin=1.6em,itemsep=2pt]
  \item \textbf{Invertibility.} The alignment map $\phi$ is constrained to be a
  bijection. A bijection cannot collapse or discard directions of the hidden space
  to manufacture a fit; it can only rotate and warp them. This is a genuine
  capacity restriction with no analogue in an unconstrained probe.
  \item \textbf{The frozen network.} The prediction is produced by the
  \emph{fixed} downstream network applied to an intervened, $\phi$-aligned hidden
  state. The map cannot directly emit a label; it can only reroute information that
  the frozen weights will then process. Spurious fits must survive this fixed
  computation.
  \item \textbf{Multi-intervention consistency.} A single $\phi$ must
  simultaneously satisfy the counterfactual constraints of \emph{several}
  intervention types (in the hierarchical equality task, the two sub-equalities
  and their conjunction). These are many coupled constraints, not one label, and
  jointly satisfying them on randomised targets is harder than fitting a single
  randomised label.
\end{enumerate}

Our working hypothesis is that these constraints make
$\rade_n^{\mathrm{DAS}}(\complexity)$ grow \emph{more slowly} than
$\rade_n^{\mathrm{DP}}(\complexity)$---i.e.\ that DAS, despite the alarming
$\complexity\to\infty$ result of \citet{sutter2025nonlinear}, may be the more
graceful failure at the finite capacities used in practice. We stress that this is
a hypothesis to be tested, not a claim; the constraints above could equally be
swamped by the sheer flexibility a non-linear $\phi$ buys, in which case DAS would
fail \emph{more} sharply. Either outcome is informative.

\section{Experimental protocol}\label{sec:protocol}

We instantiate the study on the \textbf{hierarchical equality task}, the toy
setting of \citet{sutter2025nonlinear}, for which an implementation already exists
in this repository (a $3$-layer MLP trained to decide whether two pairs of
vectors are ``both equal or both unequal'', together with linear-probe and
linear-DAS pipelines and \emph{randomised-label} controls). The task is ideal
because the ground-truth intermediate variables ($\textsc{wx}$, $\textsc{yz}$) are
known, so ``signal'' is unambiguous.

\paragraph{Complexity axes.} We add a complexity knob to each method.
For \emph{diagnostic probing}, we replace the single logistic-regression probe with
a family of MLP probes of increasing depth and width
($\complexity \in \{$linear, $1$-hidden-layer, $2$-hidden-layer, $\dots\}$), as in
\citet{pimentel2020information}. For \emph{DAS}, we replace the orthogonal
(rotation) alignment map with a family of RevNet maps of increasing depth
$L_{\mathrm{rn}}$ and hidden size $d_{\mathrm{rn}}$, as in
\citet{sutter2025nonlinear}. The data budget $n$ is held \emph{fixed} across the
sweep so that we isolate the effect of $\complexity$ (in contrast to the existing
sweeps, which vary $n$).

\paragraph{Measurements.} At each complexity level and for several seeds we record
(i) the \emph{signal}: fit on correct targets (probe accuracy / IIA on the trained
network); and (ii) the \emph{spurious capacity}
$\widehat{\rade}_n(\hypclass_{\complexity})$ of \autoref{def:capacity}, estimated
by re-fitting on randomised targets (random token labels for probing; a randomly
relabelled / random target algorithm for DAS). We additionally repeat the DAS
measurement on a \emph{randomly initialised} network, which makes the spurious
axis vivid: any IIA there is, by construction, spurious.

\paragraph{Analysis.} We (1) plot signal and spurious capacity against
$\complexity$ for each method; (2) plot the two methods together in the
\emph{signal--spurious plane} of \autoref{sec:measure}; and (3) fit a simple
growth model $\widehat{\rade}_n(\complexity) \approx a\,\complexity^{\,b}$ to
read off, and compare, the empirical growth exponents $b$ with confidence
intervals. The deliverable is a statement of the form ``per unit of spurious
capacity, method X recovers more genuine signal than method Y on this task,'' with
the supporting curves.

\paragraph{What would falsify the framing.} If signal and spurious capacity rise in
lockstep for \emph{both} methods (curves collapsing onto the diagonal of the
signal--spurious plane), then capacity carries no method-distinguishing
information and the unification, while tidy, would be practically empty. We regard
making this falsifiable as a feature.

\section{Preliminary results}\label{sec:results}

We report initial measurements on the hierarchical equality task. The signal-side
results below are stable; the spurious-capacity comparison---the report's central
quantity---comes from the scaling sweep, which we are still running.

\paragraph{WX is linearly decodable, and linearly interveneable.} A linear probe on
the trained MLP's first-layer activations recovers $\textsc{wx}$ with about $0.98$
accuracy (decodability decreases with depth; the output $O$ becomes near-perfectly
decodable only in the deeper layers, $0.70 \to 0.997$). A linear (rotation) alignment
map trained at that same layer to perform the $\textsc{wx}$ interchange reaches about
$0.999$ interchange-intervention accuracy. So, on the correct labels, both methods
recover $\textsc{wx}$, and recover it strongly: the feature is both decodable by a
probe and interveneable by a linear map.

\paragraph{An earlier negative result was a setup artefact.} Before restricting to a
single intervention, we measured DAS accuracy near chance and spent some effort on it.
That was not a property of linear alignment. The counterfactual generator produced a
multi-intervention mix in which the joint (both-subspace) intervention yielded
label-homogeneous batches---no training signal---and a single shared rotation was
being asked to satisfy several intervention objectives at once. Restricting to the
single $\textsc{wx}$ intervention (balanced batches, one objective) makes the rotation
converge cleanly: its gradient vanishes and the learned rotation stabilises as accuracy
approaches $1$. We flag this because the failure was instructive---with causal probing
it is easy to mistake a data or optimisation artefact for a substantive ``the model
does not encode this causally'' conclusion.

\paragraph{What remains.} With both methods recovering $\textsc{wx}$ on the correct
labels, the question this report is about---how readily each fits \emph{spurious}
(randomly relabelled) targets, and how that grows with complexity---is now a
meaningful comparison. That sweep (\autoref{sec:protocol}) is in progress.

\section{Implementation notes}\label{sec:limitations}

A few methodological points, mostly explaining why the setup looks the way it does.

\paragraph{Why a single intervention.} We restrict to the $\textsc{wx}$ interchange.
With the joint (both-subspace) intervention the data generator produces
label-homogeneous batches---each block fixes both intermediate values, so the output
is constant within a block---which carry no training signal; and a single shared
rotation could not, in our hands, satisfy the single and joint interventions together.
Both problems disappear with a single, balanced intervention, which is what let DAS
converge (\autoref{sec:results}).

\paragraph{Judge causal probes by held-out, order-independent evaluation.} The obvious
training-time metric---a running average of interchange accuracy---is misleading. The
counterfactual data is stored in contiguous single-type blocks, and a running average
over them drifts with the order the blocks arrive in, which can look exactly like a
model that ``learns then collapses'' even when the alignment map is frozen. Every
number we report comes from a single order-independent pass over a held-out set.

\paragraph{A pyvene gotcha.} The fast intervention path keeps only the first of
several source locations; with the (now removed) two-source joint intervention this
silently dropped the second source. With a single source it is correct and much
faster, so we use it.

\section{Relation to prior work, and a caveat on consistency}\label{sec:related}

The control-task literature is the obvious antecedent. \citet{hewitt2019control}
introduce \emph{selectivity}---task accuracy minus control-task accuracy---as a
criterion for choosing probes, and \citet{voita2020information} propose minimum
description length as a complexity-aware alternative. Our spurious-fitting capacity
is, at a single complexity level, exactly a control-task score; the novelty here is
(i) reading it as a Rademacher complexity, (ii) lifting it from diagnostic to
\emph{causal} probing, and (iii) studying its \emph{scaling} rather than its value
at one point.

We must, however, confront a tension with \citet{pimentel2020information}, which
argues \emph{against} selectivity as a probe-\emph{selection} criterion: penalising
a complex probe for its control-task accuracy artificially discards capacity that
yields a tighter information estimate. We do not retract that argument; we
repurpose the quantity. Selectivity is a poor \emph{objective} to optimise---one
should still pick the best-performing probe for estimation---but it is an excellent
\emph{descriptive diagnostic} of how much a method \emph{could} fool itself, and
its growth rate is a property of the method, not a knob to tune. Measuring
fragility is not the same as selecting against it.

Finally, a caveat on what \autoref{def:capacity} measures. It is about fitting random
targets at finite data and finite complexity. This is related to, but not the same
as, what the two theorems prove---that at infinite complexity any target can be fit.
A method could be reluctant to fit random targets at the complexities we actually
test and still be able to fit anything in the limit. Our measure speaks to the
former, the regime we use in practice, not the latter.

\section{Conclusion}\label{sec:conclusion}

We have argued that the diagnostic critique of \citet{pimentel2020information} and
the causal critique of \citet{sutter2025nonlinear} are two instances of the same
thing: a method whose reported fit grows with its capacity to fit anything, spurious
targets included. We proposed measuring that capacity directly---how well a method
fits randomly relabelled data---in a form that applies to both diagnostic probing and
DAS, and studying how it grows with complexity. Our preliminary results on the
hierarchical equality task show that, on the correct labels, the task's intermediate
variable is both linearly decodable by a probe ($\sim\!0.98$) and linearly
interveneable by an alignment map ($\sim\!0.999$ IIA); an earlier near-chance DAS
reading turned out to be an artefact of a multi-intervention data and optimisation
setup, not a property of linear alignment. With both methods recovering the signal,
the remaining and central comparison is their spurious-fitting capacity as complexity
grows. The payoff, if the sweep bears it out, is a concrete sense in which causal
probing is the safer tool at finite complexity---even though, in the limit, it is no
better than the diagnostic probing it was meant to improve on.

\small
\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
